---
output: github_document
---
git add/rm <file>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Final Project DS202
#### Marissa Baietto, Ryan Dorle, Braeden Collings

## Introduction (ANYONE CAN FIX THIS SECTION TO MAKE IT CLEANER)
* Describe topic
The goal of the project is to explore multiple datasets to better understand vehicle crashes. Understanding this topic better can lead to increased awareness, targeted policing efforts, and identification of common trends for accidents. Ultimately, the goal is the circumstances in which individuals get into vehicular accidents, and how to avoid fatal crashes.
* Why topic is important
* Research question
In pursuit of the stated goal, we will explore the following questions:

1. Does impairment affect fatality in crashes overall? When are impaired crashes most likely?

2. What regions of the United States have the most fatal crashes? What conditions are present in those regions?

3. How does the demographics of the driver affect crashes? Are changes based on occupants more prevalent for younger drivers?

4. Are crashes affected by lighting and road conditions? How so, and what conditions are most impact? 

5. When are crashes most likely? Are there any seasonal effects, and are night crashes more likely than in the morning or afternoon?

6. How is speeding related to the number of fatalities? Are younger drivers more prone to speeding? Is higher speeding limits associated with more fatalities?

## Data
### Structure
  The link to the dataset is https://www.nhtsa.gov/file-downloads?p=nhtsa/downloads/FARS/ . The NHTSA website contains a zip file containing the FARS report for every year between 1975 and 2021. Each zip file contains 33 different csv files for various aspects of a vehicle crash - data on the accident itself, the vehicles driven, damage to the vehicles, the persons in each vehicle, etc. The number of variables in these csv files ranges from 6 to 81+, and there are over 40,000 rows for each csv. 
  
  In creating a cohesive dataset, we first chose to analyze the years 2017-2021. The variables reported and elephant for the values of each variable have been through multiple changes over the years, so selecting a recent time period ensures that the reporting systems are nearly the same for each year. In addition, having a 5 year time frame to analyze allows us to see changes over time, and keeps the dataset from being so large that it is unmanageable. 
  
  Utilizing all of the csvs would have introduced too many variables, so we selected the csvs named "accident", "person", and "vehicle". "Accident" gives information on the time, place, and general aspects of the accident such as the number of fatalities and the weather and light conditions. "Person" contains data on the age and sex of each person in the crash, as well as other variables such as whether the person involved was drinking or under the influence of drugs at the time of the crash. Lastly, "vehicle" gives facts regarding the number of occupants in the vehicle, the state listed on the driver's license, what the speed limit was and whether the driver was speeding, and whether the accident was a hit and run. 
  
  The "accident" csv has separate rows for each case, while the "person" csv has separate rows for each person involved in the crash, both passengers and drivers, and the "vehicle" csv has separate rows for each vehicle involved in the crash. As seen in the code, we made decisions to limit the number of rows for each accident to ensure that the dataset would not be too large. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(readr)
library(tidyverse)
```

### Cleaning
First, we read in the 3 types of csv files for the years 2017 - 2021. 
```{r, warning=FALSE,message=FALSE}
accident_2017 <- read_csv("2017.csv")
accident_2018 <- read_csv("2018.csv")
accident_2019 <- read_csv("2019.csv")
accident_2020 <- read_csv("2020.csv")
accident_2021 <- read_csv("2021.csv")

person_2017 <- read_csv("2017_person.csv")
person_2018 <- read_csv("2018_person.csv")
person_2019 <- read_csv("2019_person.csv")
person_2020 <- read_csv("2020_person.csv")
person_2021 <- read_csv("2021_person.csv")

vehicle_2017 <- read_csv("2017_vehicle.csv")
vehicle_2018 <- read_csv("2018_vehicle.csv")
vehicle_2019 <- read_csv("2019_vehicle.csv")
vehicle_2020 <- read_csv("2020_vehicle.csv")
vehicle_2021 <- read_csv("2021_vehicle.csv")
```

Originally, the "person" csv files were 24.1 MB large for each year. GitHub's size limit for uploading files to the reposirory is 24MB, which meant that we could not upload the "person" csvs to GitHub. To overcome this problem, we created a separate R Script and slimmed the csv files there. We did so by opening the csv files for each year and filtering such that (PER_NO == 1), meaning that only the driver is included in the csv files. This brought the size of the csv down to meet the limits. We then saved the new csv files as "(YEAR)_person.csv", and uploaded the modified csv files to this project report. 

Then, we selected the variables we needed for each type of csv.
```{r, warning=FALSE,message=FALSE}
accident_2017 <- accident_2017 %>% select(STATENAME, ST_CASE, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, LGT_CONDNAME, WEATHER1NAME, FATALS, DRUNK_DR, DAY_WEEK)
accident_2018 <- accident_2018 %>% select(STATENAME, ST_CASE, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, LGT_CONDNAME, WEATHER1NAME, FATALS, DRUNK_DR, DAY_WEEK)
accident_2019 <- accident_2019 %>% select(STATENAME, ST_CASE, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, LGT_CONDNAME, WEATHER1NAME, FATALS, DRUNK_DR, DAY_WEEK)
accident_2020 <- accident_2020 %>% select(STATENAME, ST_CASE, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, LGT_CONDNAME, WEATHERNAME, FATALS, DRUNK_DR, DAY_WEEK)
accident_2021 <- accident_2021 %>% select(STATENAME, ST_CASE, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, LGT_CONDNAME, WEATHERNAME, FATALS, DAY_WEEK)

person_2017 <- person_2017 %>% select(AGE, SEXNAME, DRINKINGNAME, DRUGSNAME, LAG_HRSNAME, ST_CASE, VEH_NO, PER_NO)
person_2018 <- person_2018 %>% select(AGE, SEXNAME, DRINKINGNAME, DRUGSNAME, LAG_HRSNAME, ST_CASE, VEH_NO, PER_NO)
person_2019 <- person_2019 %>% select(AGE, SEXNAME, DRINKINGNAME, DRUGSNAME, LAG_HRSNAME, ST_CASE, VEH_NO, PER_NO)
person_2020 <- person_2020 %>% select(AGE, SEXNAME, DRINKINGNAME, DRUGSNAME, LAG_HRSNAME, ST_CASE, VEH_NO, PER_NO)
person_2021 <- person_2021 %>% select(AGE, SEXNAME, DRINKINGNAME, DRUGSNAME, LAG_HRSNAME, ST_CASE, VEH_NO, PER_NO)
```

There were 2 disreptencies between the years in the accident files. First, "accident_2021" did not have a variable for "DRUNK_DR". Second, the naming of the variable containing information on the weather changed in 2020 and 2021 from "WEATHER1NAME" to "WEATHERNAME". (THIS WAS A GRAPH THAT DIDN"T MAKE SENSE IN RYAN"S SECTION) We accounted for the disreptency in weather by creating a duplicate variable named "WEATHERNAME" in the 2017,2018,2019 dataframes and deleting the "WEATHER1NAME" variables in those same dataframes. 
```{r}
accident_2017$WEATHERNAME <- accident_2017$WEATHER1NAME
accident_2018$WEATHERNAME <- accident_2018$WEATHER1NAME
accident_2019$WEATHERNAME <- accident_2019$WEATHER1NAME

accident_2017 <- accident_2017 %>% select(-WEATHER1NAME)
accident_2018 <- accident_2018 %>% select(-WEATHER1NAME)
accident_2019 <- accident_2019 %>% select(-WEATHER1NAME)
```

To enable us to merge the datasets for each year, we created a "YEAR" variable for the "person" and "vehicle" datasets.
```{r, warning=FALSE,message=FALSE}
person_2017$YEAR = 2017
person_2018$YEAR = 2018
person_2019$YEAR = 2019
person_2020$YEAR = 2020
person_2021$YEAR = 2021

vehicle_2017$YEAR = 2017
vehicle_2018$YEAR = 2018
vehicle_2019$YEAR = 2019
vehicle_2020$YEAR = 2020
vehicle_2021$YEAR = 2021
```

To merge the datasets together, we first created 5 dataframes for each year, with each dataframe being a merge of the year's accident, person, and year dataframes. This was performed with a left_join with "accident" as the main dataframe. After "accident" and "person" were merged by year and st_case. 

ST_CASE is a unique identifier for each accident that occured in the state. Vehicles and persons involved in the same crash have identical ST_CASE numbers. For each state, ST_CASE starts with a new number, but has the same format. For example, Alabama's ST_CASE numbers are in the 10,000s, Alaska's ST_CASE numbers are in the 20,000s, and so on with the states in order alphabetically. ST_CASE numbers reset each year, meaning that the first case in Alabama in 2017, 2018, 2019, 2020, and 2021 have the ST_CASE number 10001.

We merged the new dataframes with the vehicle dataframes for each year by performing a left merge by year, st_case, and veh_no. VEH_NO is a unique identifier for each vehicle involved in the crash, and it resets to 1 for each new accident. Merging on VEH_No ensured that we were matching each driver to the vehicle they drove, as the "person" file included the vehicle number as well for each individual.
```{r, warning=FALSE,message=FALSE}
df_2017 <- left_join(accident_2017, person_2017, by=c('YEAR', 'ST_CASE'))
df_2018 <- left_join(accident_2018, person_2018, by=c('YEAR', 'ST_CASE'))
df_2019 <- left_join(accident_2019, person_2019, by=c('YEAR', 'ST_CASE'))
df_2020 <- left_join(accident_2020, person_2020, by=c('YEAR', 'ST_CASE'))
df_2021 <- left_join(accident_2021, person_2021, by=c('YEAR', 'ST_CASE'))


df_2017 <- left_join(df_2017, vehicle_2017, by=c('YEAR', 'ST_CASE', 'VEH_NO'))
df_2018 <- left_join(df_2018, vehicle_2018, by=c('YEAR', 'ST_CASE', 'VEH_NO'))
df_2019 <- left_join(df_2019, vehicle_2019, by=c('YEAR', 'ST_CASE', 'VEH_NO'))
df_2020 <- left_join(df_2020, vehicle_2020, by=c('YEAR', 'ST_CASE', 'VEH_NO'))
df_2021 <- left_join(df_2021, vehicle_2021, by=c('YEAR', 'ST_CASE', 'VEH_NO'))
```

To merge all of the years together, we used the bind_rows function and put the years in order from 2017 to 2021. Again, we confronted the problem that the file was too large to upload to GitHub. So, we selected one vehicle per crash, by filtering such that VEH_NO == 1. There is always a vehicle with number equal to 1 for each accident because the first vehicle listed for each accident has a vehicle number equal to one. We also only chose the even state case numbers. This ensured that each state would have the same proportion of accidents that they had in the original file, and choosing odd state case numbers gave us a smaller random sample from the original dataset. 
```{r, warning=FALSE,message=FALSE}
df <- bind_rows(df_2017, df_2018,df_2019,df_2020,df_2021)
df1 <- df %>% group_by(STATENAME, YEAR) %>% filter(VEH_NO == 1 & ST_CASE %% 2 == 0) #add & ST_CASE %% 2 == 0
```

We noticed that there were 2 strange columns in the final dataframe named '...22' and '...21'. We believe that these were created when we joined the 'accident', 'person', and 'vehicle' datasets, as there were so many variables that R was not able to show them all when we viewed the merged datasets. We took these columns out, as they had no value to our analysis. 
```{r, warning=FALSE,message=FALSE}
colnames(df1)
df1 <- df1 %>% select(-one_of('...22', '...21'))
```

To ease running the code, we saved the final dataset to "master.csv" so that we could then only run the line importing "master.csv" rather than run the code above repeatedly.
```{r, warning=FALSE,message=FALSE}
write.csv(df1, 'master.csv')
```

### Variables
* STATENAME: The state the fatal crash occured in
* WEATHERNAME: The type of weather outside at the time and location of the crash.
* VEH_NO: Vehicle Number, a number assigned to each vehicle in the crash.
* PER_NO: Person Number, A unique number assigned to any person involved in the crash.
* DRUGSNAME: Were there drugs involved in the crash if so the name of the drug.
* DRINKINGNAME: Was drinking involved in the crash if so what alcohol.       
* SEXNAME: The gender of people related to the crash.
* AGE: Age of people in the crash.
* DRUNK_DR: Drinking Drivers, The number of drinking drivers involved in the crash.
* FATALS: The number of fatalities in the crash.
* WEATHER1NAME: Atmospheric conditions at the time of the crash. 
* LGT_CONDNAME: What was the lighting condition at the time of the crash.
* LATITUDE: The location of the crash using global positioning (East and West).
* LONGITUD: The location of the crash using global positioning (North and South).
* RUR_URBNAME: Land Use, the segment of the traffic way on which the crash occurred based on urbanized areas.
* HOUR: The time of the crash in hours.
* MINUTE: The time of the crash in minutes.
* MONTH: The month the crash occurred in.
* DAY: The day the crash occurred on.
* YEAR: The year in which the crash happened. 
* ST_CASE: Unique case number assigned to each crash.
*	HIT_RUN: Whether the crash was a hit and run.
*	NUMOCCS: Number of occupants in the vehicle.
*	L_STATE: The state listed on the driver’s license. 
*	SPEEDREL: Whether the crash was speeding related.
*	VSPD_LIM: What the speeding limit was.

## Methods
* Methods utilized, if used

## Results
* All results and graphics
    + For each graphic, 
        - Motivation for making the graphic
        - Insights obtained from the graphics
        - Necessary explanations
    + Formatting
        - Axis Labels, Title, Captions, Color Scheme if necessary
        - Each graphic illustrates one point
        - Make color scheme cohesive
        - Avoid repetition
* List trials and errors
    + Critical of findings
    + Multiple approaches and techniques used to verify unintuitive results
* Any additional research used to help understand/explain findings

```{r, warning=FALSE,message=FALSE}
master <- read_csv('master.csv')
head(master)
```

### When are crashes most likely? Are there any seasonal effects, and are night crashes more likely than in the morning or afternoon?

```{r}
library(tidyverse)
master %>% group_by(YEAR) %>% ggplot(aes(x = YEAR, weight = FATALS)) + geom_bar() + ggtitle("Fatalities Over Last 5 Years") + ylab("Fatalities") + xlab("Year")
```
```{r}
df %>% group_by(MONTH) %>% summarise(count = sum(FATALS)) %>% arrange(desc(count))
```

```{r}
farb <- c("#4100f3", "#1e00ff", "#5700e7", "#6900d9", "#dd00de", "#d3007f", "#c80028", "#ce0053", "#d900ae", "#c91e1e", "#b500e4", "#8a00e9")
farb<-c("#c91e1e","#c80028","#ce0053","#d3007f","#d900ae","#dd00de","#b500e4", "#8a00e9","#6900d9","#5700e7","#4100f3", "#1e00ff")
plot_names <- c('1' = "January",
                '2' = "February",
                '3' = "March",
                '4' = "April",
                '5' = "May",
                '6' = "June",
                '7' = "July",
                '8' = "August",
                '9' = "September",
                '10' = "October",
                '11' = "November",
                '12' = "December")

master %>% group_by(DAY) %>% ggplot(aes(x = DAY, weight = FATALS)) + geom_bar(aes(fill = factor(MONTH, levels = c("10","7","8","6","9","5","11","12","4","3","1","2"))))+ facet_wrap(~MONTH, labeller = as_labeller(plot_names)) +geom_hline(yintercept = 200, color='RED') + theme(legend.position="bottom") + scale_fill_manual(values= farb)+ theme( legend.position="right")+labs(fill='Monthly Fatalities') + ylab("Fatalities") + ggtitle("Fatalities by Day and Month")
```
Originally showed same pattern for every month.
Originally showed high in January and Feburary and low for rest of year.


```{r}
df1 %>% ggplot(aes(x = MONTH)) + geom_histogram()
```

It appears that fatal crashes occur the most in the beginning of the year, and decrease heavily until December for all of the years.

```{r}
df1 %>% ggplot(aes(x = HOUR)) + geom_histogram(bins = 100) + ylab("Fatalities") + ggtitle("Fatalities by Hour")
df1 %>% filter(HOUR >24)
df1 %>% filter(HOUR <= 24) %>% ggplot(aes(x = as.factor(HOUR), weight = FATALS)) + geom_bar() + ylab("Fatalities") + ggtitle("Fatalities by Hour") + xlab("HOUR")
```

```{r}
summary(df1$FATALS)
```



```{r}
df1 %>% ggplot(aes(x = as.factor(DAY_WEEK), weight = FATALS)) + geom_bar() + scale_x_discrete(
                      labels=c("Sunday","Monday", "Tuesday", "Wednesday","Thursday","Friday", "Saturday")) + xlab("Weekday") + ylab("Fatalities") + ggtitle("Fatalities by Weekday")
```

There is a strange outlier with hour, so I will remove that.

```{r}
df1 %>% filter(HOUR <=24) %>% ggplot(aes(x = HOUR)) + geom_histogram() + scale_x_continuous(name="Hour", limits=c(0, 24))
```
It appears that the peak is around 17:00 military time, which is 5pm. Thus, crashes appear to be increasingly likely as the day goes on, peaks at 5pm, and has a minimum at around 4pm.

### How is speeding related to the number of fatalities? Are younger drivers more prone to speeding? Is higher speeding limits associated with more fatalities?



### Does impairment affect fatality in crashes overall? When are impaired crashes most likely?

```{r}
master$ImpairmentAlcohol <- ifelse(master$DRINKINGNAME %in% c("No (Alcohol Not Involved)", "Unknown (Police Reported)", "Not Reported", "Reported as Unknown"), FALSE, TRUE)

master$ImpairmentDrugs <- ifelse(master$DRUGSNAME %in% c("No (drugs not involved)", "Unknown (Police Reported)", "Not Reported", "Reported as Unknown"), FALSE, TRUE)

master <- master %>% mutate(
  Impairment = as.logical(pmax(ImpairmentDrugs, ImpairmentAlcohol))
)

master %>% ggplot(aes(x = Impairment)) + geom_bar()

```

```{r}
master %>% filter(Impairment == TRUE) %>%
  filter(HOUR <=24) %>% ggplot(aes(x = HOUR), fill = factor(MONTH)) + geom_bar(bins = 24)
```





### What regions of the United States have the most fatal crashes? What conditions are present in those regions?


```{r}
master <- master %>% 

  mutate(Region = case_when( 

    STATENAME %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "District of Columbia", "New Jersey", "New York", "Pennsylvania") ~ "Northeast", 

    STATENAME %in% c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota", "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin") ~ "Midwest", 

    STATENAME %in% c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia") ~ "South", 

    STATENAME %in% c("Alaska", "Arizona", "California", "Colorado", "Hawaii", "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming") ~ "West", 

    TRUE ~ "Other" 

  )) 

master %>% ggplot(aes(x = Region)) + geom_bar()

```
```{r}
head(master)

```

## Conclusion
* Overall idea
* Interesting further questions
* Ideas for future research

