---
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Final Project DS202
## Marissa Baietto, Ryan Dorle, Braeden Collings
Importing the accident datasets.
```{r}
library(readr)
accident_2017 <- read_csv("2017.csv")
accident_2018 <- read_csv("2018.csv")
accident_2019 <- read_csv("2019.csv")
accident_2020 <- read_csv("2020.csv")
accident_2021 <- read_csv("2021.csv")
```

```{r}
library(tidyverse)
```

Selecting the variables we need.
```{r}
accident_2017 <- accident_2017 %>% select(STATENAME, ST_CASE, PERNOTMVIT, VE_FORMS, PVH_INVL, PERMVIT, COUNTY, CITY, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, HARM_EVNAME, MAN_COLLNAME, RELJCT2NAME, WRK_ZONENAME, LGT_CONDNAME, WEATHER1NAME, SCH_BUSNAME, FATALS, DRUNK_DR)
accident_2018 <- accident_2018 %>% select(STATENAME, ST_CASE, PERNOTMVIT, VE_FORMS, PVH_INVL, PERMVIT, COUNTY, CITY, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, HARM_EVNAME, MAN_COLLNAME, RELJCT2NAME, WRK_ZONENAME, LGT_CONDNAME, WEATHER1NAME, SCH_BUSNAME, FATALS, DRUNK_DR)
accident_2019 <- accident_2019 %>% select(STATENAME, ST_CASE, PERNOTMVIT, VE_FORMS, PVH_INVL, PERMVIT, COUNTY, CITY, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, HARM_EVNAME, MAN_COLLNAME, RELJCT2NAME, WRK_ZONENAME, LGT_CONDNAME, WEATHER1NAME, SCH_BUSNAME, FATALS, DRUNK_DR)
accident_2020 <- accident_2020 %>% select(STATENAME, ST_CASE, PERNOTMVIT, VE_FORMS, PVH_INVL, PERMVIT, COUNTY, CITY, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, HARM_EVNAME, MAN_COLLNAME, RELJCT2NAME, WRK_ZONENAME, LGT_CONDNAME, WEATHERNAME, SCH_BUSNAME, FATALS, DRUNK_DR)
accident_2021 <- accident_2021 %>% select(STATENAME, ST_CASE, PERNOTMVIT, VE_FORMS, PVH_INVL, PERMVIT, COUNTY, CITY, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, HARM_EVNAME, MAN_COLLNAME, RELJCT2NAME, WRK_ZONENAME, LGT_CONDNAME, WEATHERNAME, SCH_BUSNAME, FATALS)
```

Importing the person datasets.
```{r}
person_2017 <- read_csv("2017_person.csv")
person_2018 <- read_csv("2018_person.csv")
person_2019 <- read_csv("2019_person.csv")
person_2020 <- read_csv("2020_person.csv")
person_2021 <- read_csv("2021_person.csv")
```

Adding year variable to the datasets.
```{r}
person_2017$YEAR = 2017
person_2018$YEAR = 2018
person_2019$YEAR = 2019
person_2020$YEAR = 2020
person_2021$YEAR = 2021
```


Merging the datasets.
```{r}
df_2017 <- left_join(accident_2017, person_2017, by=c('YEAR', 'ST_CASE'))
df_2018 <- left_join(accident_2018, person_2018, by=c('YEAR', 'ST_CASE'))
df_2019 <- left_join(accident_2019, person_2019, by=c('YEAR', 'ST_CASE'))
df_2020 <- left_join(accident_2020, person_2020, by=c('YEAR', 'ST_CASE'))
df_2021 <- left_join(accident_2021, person_2021, by=c('YEAR', 'ST_CASE'))
```

Creating one dataset with all years.
```{r}
df <- bind_rows(df_2017, df_2018,df_2019,df_2020,df_2021)
df1 <- df %>% group_by(STATENAME, YEAR) %>% filter(ST_CASE %% 10000 <= 200)
```

Saving the final dataset to 'master.csv'

```{r}
write.csv(df1, 'master.csv')
```

## Discription of Data
The data we chose to work with is from the National Highway Traffic Safety Administration. The administration has a system called FARS or the Fatality Analysis Reporting System in which vehicle crashes are noted. The data is comprised from police reports, death certificates, various medical reports, and department data. We chose to look at the years from 2017-2021 for our report. 
```{r}
head(df1)

```

## Marginal Summaries
* STATENAME: The state the fatal crash occured in
* WEATHERNAME: The type of weather outside at the time and location of the crash.
* VEH_NO: Vehicle Number, a number assigned to each vehicle in the crash.
* PER_NO: Person Number, A unique number assigned to any person involved in the crash.
* WORK_INJNAME: Fatal injury at work, whether the person was at work at the time of the crash.
* LAG_HRSNAME: Lag Time, hours between the crash and the time of death.
* DOANAME:
* DRUGSNAME: Were there drugs involved in the crash if so the name of the drug.
* DRINKINGNAME: Was drinking involved in the crash if so what alcohol.       
* EJECTIONNAME: Was there an ejection of a person in the crash.
* REST_USENAME: Restraint System Used, was there a seatbelt used in the crash.
* INJ_SEVNAME: What is the severity of the injury. 
* SEXNAME: The gender of people related to the crash.
* AGE: Age of people in the crash.
* DRUNK_DR: Drinking Drivers, The number of drinking drivers involved in the crash.
* FATALS: The number of fatalities in the crash.
* SCH_BUSNAME: Was there a school bus related in the crash.
* WEATHER1NAME: Atmospheric conditions at the time of the crash. 
* LGT_CONDNAME: What was the lighting condition at the time of the crash.
* WRK_ZONENAME: Did the crash occur in the boundaries of a work zone.      
* RELJCT2NAME: Location of the crash with respect to presence to a junction or interchange areas.     
* MAN_COLLNAME: Manner of the collision, were ther multiple vechiles involed in the crash. 
* HARM_EVNAME: First harmful event, first injury produced by the event of a crash. 
* LATITUDE: The location of the crash using global positioning (East and West).
* LONGITUD: The location of the crash using global positioning (North and South).
* RUR_URBNAME: Land Use, the segment of the traffic way on which the crash occurred based on urbanized areas.
* HOUR: The time of the crash in hours.
* MINUTE: The time of the crash in minutes.
* CITY: The location of the crash.
* MONTH: The month the crash occurred in.
* DAY: The day the crash occurred on.
* YEAR: The year in which the crash happened. 
* COUNTY: The county in which the crash occurred in.
* PERMVIT: Number of motorists involved in the crash.
* PVH_INVL: Number of parked vehicles involved in the crash.
* VE_FORMS: Number of vehicles in-transport involved in the crash.
* PERNOTMVIT: The number of non-motorists in the crash.
* ST_CASE: Unique case number assigned to each crash.

# Research Goals and Questions

The goal of the project is to explore the data set to better understand vehicle crashes. Understanding this topic better can lead to increased awareness, targeted policing efforts, and identification of common trends for accidents. Ultimately, the goal is the circumstances in which individuals get into vehicular accidents, and how to avoid fatal crashes.

In pursuit of the stated goal, we will explore the following questions:

1. Does impairment affect crashes overall? When are impaired crashes most likely?

2. What regions of the United States have the most fatal crashes? What conditions are present in those regions?

3. How does the demographics of the driver affect crashes? Are changes based on occupants more prevalent for younger drivers?

4. Are crashes affected by lighting and road conditions? How so, and what conditions are most impact? Sent from my iPhone
