---
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Final Project DS202
## Marissa Baietto, Ryan Dorle, Braeden Collings

Here is the link to all of the csv files we used: https://www.nhtsa.gov/file-downloads?p=nhtsa/downloads/FARS/

Importing the accident datasets.
```{r}
library(readr)
accident_2017 <- read_csv("2017.csv")
accident_2018 <- read_csv("2018.csv")
accident_2019 <- read_csv("2019.csv")
accident_2020 <- read_csv("2020.csv")
accident_2021 <- read_csv("2021.csv")
```

```{r}
library(tidyverse)
```

Selecting the variables we need.
```{r}
accident_2017 <- accident_2017 %>% select(STATENAME, ST_CASE, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, LGT_CONDNAME, WEATHER1NAME, FATALS, DRUNK_DR, DAY_WEEK)
accident_2018 <- accident_2018 %>% select(STATENAME, ST_CASE, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, LGT_CONDNAME, WEATHER1NAME, FATALS, DRUNK_DR, DAY_WEEK)
accident_2019 <- accident_2019 %>% select(STATENAME, ST_CASE, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, LGT_CONDNAME, WEATHER1NAME, FATALS, DRUNK_DR, DAY_WEEK)
accident_2020 <- accident_2020 %>% select(STATENAME, ST_CASE, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, LGT_CONDNAME, WEATHERNAME, FATALS, DRUNK_DR, DAY_WEEK)
accident_2021 <- accident_2021 %>% select(STATENAME, ST_CASE, MONTH, DAY, YEAR, HOUR, MINUTE, RUR_URBNAME, LATITUDE, LONGITUD, LGT_CONDNAME, WEATHERNAME, FATALS, DAY_WEEK)
#2021 does not have drunk driving variable

accident_2017$WEATHERNAME <- accident_2017$WEATHER1NAME
accident_2018$WEATHERNAME <- accident_2018$WEATHER1NAME
accident_2019$WEATHERNAME <- accident_2019$WEATHER1NAME

accident_2017 <- accident_2017 %>% select(-WEATHER1NAME)
accident_2018 <- accident_2018 %>% select(-WEATHER1NAME)
accident_2019 <- accident_2019 %>% select(-WEATHER1NAME)
```


Importing the person datasets.
```{r}
person_2017 <- read_csv("2017_person.csv")
person_2018 <- read_csv("2018_person.csv")
person_2019 <- read_csv("2019_person.csv")
person_2020 <- read_csv("2020_person.csv")
person_2021 <- read_csv("2021_person.csv")
```

Selecting person variables.
```{r}
person_2017 <- person_2017 %>% select(AGE, SEXNAME, DRINKINGNAME, DRUGSNAME, LAG_HRSNAME, ST_CASE, VEH_NO, PER_NO)
person_2018 <- person_2018 %>% select(AGE, SEXNAME, DRINKINGNAME, DRUGSNAME, LAG_HRSNAME, ST_CASE, VEH_NO, PER_NO)
person_2019 <- person_2019 %>% select(AGE, SEXNAME, DRINKINGNAME, DRUGSNAME, LAG_HRSNAME, ST_CASE, VEH_NO, PER_NO)
person_2020 <- person_2020 %>% select(AGE, SEXNAME, DRINKINGNAME, DRUGSNAME, LAG_HRSNAME, ST_CASE, VEH_NO, PER_NO)
person_2021 <- person_2021 %>% select(AGE, SEXNAME, DRINKINGNAME, DRUGSNAME, LAG_HRSNAME, ST_CASE, VEH_NO, PER_NO)
```


Adding year variable to the datasets.
```{r}
person_2017$YEAR = 2017
person_2018$YEAR = 2018
person_2019$YEAR = 2019
person_2020$YEAR = 2020
person_2021$YEAR = 2021
```

Importing Vehicle Datasets
```{r}
vehicle_2017 <- read_csv("2017_vehicle.csv")
vehicle_2018 <- read_csv("2018_vehicle.csv")
vehicle_2019 <- read_csv("2019_vehicle.csv")
vehicle_2020 <- read_csv("2020_vehicle.csv")
vehicle_2021 <- read_csv("2021_vehicle.csv")
```

Adding year variable to the datasets.
```{r}
vehicle_2017$YEAR = 2017
vehicle_2018$YEAR = 2018
vehicle_2019$YEAR = 2019
vehicle_2020$YEAR = 2020
vehicle_2021$YEAR = 2021
```

Merging the datasets.
```{r}
df_2017 <- left_join(accident_2017, person_2017, by=c('YEAR', 'ST_CASE'))
df_2018 <- left_join(accident_2018, person_2018, by=c('YEAR', 'ST_CASE'))
df_2019 <- left_join(accident_2019, person_2019, by=c('YEAR', 'ST_CASE'))
df_2020 <- left_join(accident_2020, person_2020, by=c('YEAR', 'ST_CASE'))
df_2021 <- left_join(accident_2021, person_2021, by=c('YEAR', 'ST_CASE'))


df_2017 <- left_join(df_2017, vehicle_2017, by=c('YEAR', 'ST_CASE', 'VEH_NO'))
df_2018 <- left_join(df_2018, vehicle_2018, by=c('YEAR', 'ST_CASE', 'VEH_NO'))
df_2019 <- left_join(df_2019, vehicle_2019, by=c('YEAR', 'ST_CASE', 'VEH_NO'))
df_2020 <- left_join(df_2020, vehicle_2020, by=c('YEAR', 'ST_CASE', 'VEH_NO'))
df_2021 <- left_join(df_2021, vehicle_2021, by=c('YEAR', 'ST_CASE', 'VEH_NO'))
```

Creating one dataset with all years.
```{r}
df <- bind_rows(df_2017, df_2018,df_2019,df_2020,df_2021)
df1 <- df %>% group_by(STATENAME, YEAR) %>% filter(VEH_NO == 1 & ST_CASE %% 2 == 0) #add & ST_CASE %% 2 == 0
```
```{r}
colnames(df1)
df1 <- df1 %>% select(-one_of('...22', '...21'))
```


Saving the final dataset to 'master.csv'

```{r}
write.csv(df1, 'master.csv')
```


## Discription of Data
The data we chose to work with is from the National Highway Traffic Safety Administration. The administration has a system called FARS or the Fatality Analysis Reporting System in which vehicle crashes are noted. The data is comprised from police reports, death certificates, various medical reports, and department data. We chose to look at the years from 2017-2021 for our report. 
```{r}
master <- read_csv('master.csv')
head(master)

```

## Marginal Summaries
* STATENAME: The state the fatal crash occured in
* WEATHERNAME: The type of weather outside at the time and location of the crash.
* VEH_NO: Vehicle Number, a number assigned to each vehicle in the crash.
* PER_NO: Person Number, A unique number assigned to any person involved in the crash.
* DRUGSNAME: Were there drugs involved in the crash if so the name of the drug.
* DRINKINGNAME: Was drinking involved in the crash if so what alcohol.       
* SEXNAME: The gender of people related to the crash.
* AGE: Age of people in the crash.
* DRUNK_DR: Drinking Drivers, The number of drinking drivers involved in the crash.
* FATALS: The number of fatalities in the crash.
* WEATHER1NAME: Atmospheric conditions at the time of the crash. 
* LGT_CONDNAME: What was the lighting condition at the time of the crash.
* LATITUDE: The location of the crash using global positioning (East and West).
* LONGITUD: The location of the crash using global positioning (North and South).
* RUR_URBNAME: Land Use, the segment of the traffic way on which the crash occurred based on urbanized areas.
* HOUR: The time of the crash in hours.
* MINUTE: The time of the crash in minutes.
* MONTH: The month the crash occurred in.
* DAY: The day the crash occurred on.
* YEAR: The year in which the crash happened. 
* ST_CASE: Unique case number assigned to each crash.
*	HIT_RUN: Whether the crash was a hit and run.
*	NUMOCCS: Number of occupants in the vehicle.
*	L_STATE: The state listed on the driver’s license. 
*	SPEEDREL: Whether the crash was speeding related.
*	VSPD_LIM: What the speeding limit was.


# Research Goals and Questions

The goal of the project is to explore the data set to better understand vehicle crashes. Understanding this topic better can lead to increased awareness, targeted policing efforts, and identification of common trends for accidents. Ultimately, the goal is the circumstances in which individuals get into vehicular accidents, and how to avoid fatal crashes.

In pursuit of the stated goal, we will explore the following questions:

1. Does impairment affect fatality in crashes overall? When are impaired crashes most likely?

2. What regions of the United States have the most fatal crashes? What conditions are present in those regions?

3. How does the demographics of the driver affect crashes? Are changes based on occupants more prevalent for younger drivers?

4. Are crashes affected by lighting and road conditions? How so, and what conditions are most impact? 

5. When are crashes most likely? Are there any seasonal effects, and are night crashes more likely than in the morning or afternoon?

6. How is speeding related to the number of fatalities? Are younger drivers more prone to speeding? Is higher speeding limits associated with more fatalities?


### When are crashes most likely? Are there any seasonal effects, and are night crashes more likely than in the morning or afternoon?

```{r}
master %>% group_by(YEAR) %>% ggplot(aes(x = YEAR, weight = FATALS)) + geom_bar() + ggtitle("Fatalities Over Last 5 Years") + ylab("Fatalities") + xlab("Year")
```
```{r}
df %>% group_by(MONTH) %>% summarise(count = sum(FATALS)) %>% arrange(desc(count))
```


```{r}
farb <- c("#4100f3", "#1e00ff", "#5700e7", "#6900d9", "#9100ac", "#a70085", "#ba004d", "#b1006d", "#9d009a", "#c30004", "#8500bc", "#7800cb")
farb <- c("#4100f3", "#1e00ff", "#5700e7", "#6900d9", "#dd00de", "#d3007f", "#c80028", "#ce0053", "#d900ae", "#c91e1e", "#b500e4", "#8a00e9")

plot_names <- c('1' = "January",
                '2' = "February",
                '3' = "March",
                '4' = "April",
                '5' = "May",
                '6' = "June",
                '7' = "July",
                '8' = "August",
                '9' = "September",
                '10' = "October",
                '11' = "November",
                '12' = "December")

master %>% group_by(DAY) %>% ggplot(aes(x = DAY, weight = FATALS)) + geom_bar(aes(fill = as.factor(MONTH))) + facet_wrap(~MONTH, labeller = as_labeller(plot_names))+geom_hline(yintercept = 200, color='RED') + theme(legend.position="bottom") + scale_fill_manual(values= farb)
```

```{r}
df1 %>% ggplot(aes(x = MONTH)) + geom_histogram()
```

It appears that fatal crashes occur the most in the beginning of the year, and decrease heavily until December for all of the years.

```{r}
df1 %>% ggplot(aes(x = HOUR)) + geom_histogram(bins = 100)
```

```{r}
summary(df1$FATALS)
```



```{r}
df1 %>% ggplot(aes(x = as.factor(DAY_WEEK), weight = FATALS)) + geom_bar() + scale_x_discrete(
                      labels=c("Sunday","Monday", "Tuesday", "Wednesday","Thursday","Friday", "Saturday")) 
```

There is a strange outlier with hour, so I will remove that.

```{r}
df1 %>% filter(HOUR <=24) %>% ggplot(aes(x = HOUR)) + geom_histogram() + scale_x_continuous(name="Hour", limits=c(0, 24))
```
It appears that the peak is around 17:00 military time, which is 5pm. Thus, crashes appear to be increasingly likely as the day goes on, peaks at 5pm, and has a minimum at around 4pm.

### How is speeding related to the number of fatalities? Are younger drivers more prone to speeding? Is higher speeding limits associated with more fatalities?



### Does impairment affect fatality in crashes overall? When are impaired crashes most likely?

```{r}
master$ImpairmentAlcohol <- ifelse(master$DRINKINGNAME %in% c("No (Alcohol Not Involved)", "Unknown (Police Reported)", "Not Reported", "Reported as Unknown"), FALSE, TRUE)

master$ImpairmentDrugs <- ifelse(master$DRUGSNAME %in% c("No (drugs not involved)", "Unknown (Police Reported)", "Not Reported", "Reported as Unknown"), FALSE, TRUE)

master <- master %>% mutate(
  Impairment = as.logical(pmax(ImpairmentDrugs, ImpairmentAlcohol))
)

master %>% ggplot(aes(x = Impairment)) + geom_bar()

```

```{r}
master %>% filter(Impairment == TRUE) %>%
  filter(HOUR <=24) %>% ggplot(aes(x = HOUR), fill = factor(MONTH)) + geom_bar(bins = 24)
```





### What regions of the United States have the most fatal crashes? What conditions are present in those regions?

```{r}
master$West <- ifelse(master$STATENAME %in% c("Alaska", "Arizona", "California","Colorado", "Hawaii", "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming"), "West", NA)

  
master$NorthEast <- ifelse(master$STATENAME %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New Jersey", "New York", "Pennsylvania"), "NorthEast", NA)

master$South <- ifelse(master$STATENAME %in% c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia"), "South", NA)

master$MidWest <- ifelse(master$STATENAME %in% c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota", "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin"), "MidWest", NA)

master <- master %>% mutate(
  Region = as.logical(pmax(NorthEast, West, South, MidWest))
)

```



